# A minimal example

```{r}
#| results: "asis"
#| echo: false
source("_common.R")
ans <- append_df(refresh = TRUE)
status("drafting", "2022-12-21", section = "A minimal example")  # drafting, polishing, complete
```

## Installation

If you use R installing X-13ARIMA-SEATS from CRAN is as easy as installing any other R package [@jss]:

``` r
install.packages("seasonal")
```

## A minimal example

Once the package is installed it can be loaded in the usual way:

```{r}
library(seasonal)
```

The `seas()` function provides the core functionality of the package.
By default, `seas` calls the automatic procedures of X-13 to perform a seasonal adjustment that works well in most circumstances:

```{r}
seas(unemp)
```

The first argument of seas is a time series of class "ts".
The unemp example series measures US unemployment and is included in seasonal.
The function returns an object of class "seas" that contains the necessary information on the adjustment performed on this time series.

```{r}
m <- seas(unemp)
```

There are several functions and methods for "seas" objects.
The final function returns the adjusted series.
The plot method shows a plot with the unadjusted and the adjusted series.

```{r}
plot(m)
```

The summary method displays an overview of the model, very similar to the one produced by R's lm function:

```{r}
summary(m)
```
