# How to use the book

```{r}
#| results: "asis"
#| echo: false
source("_common.R")
status("drafting", "2022-12-21", section = "How to use the book")  # drafting, polishing, complete
```

-   Overview of the book

-   The Book wants to give concrete advise in case of a problem.

-   Ideally we want to have a quick check list that gives readers a starting point where to look for further advice.
    That could be something like a Cheat Sheet (https://www.rstudio.com/resources/cheatsheets/) with quick advise and chapter references.

-   For each section, we want to provide a concrete and informative case study.
    Some examples are provided in the outline.

This text will focus on seasonal adjustment and its implementation in R.

Specifically, the audience will be both R users who want to learn about seasonal adjustment as well as seasonal adjustment practitioners, who are interested in using R.
The book will be tailored to the practical applications of seasonal adjustment within R.
It presents background material and references for the theoretically minded reader.
The main focus, however, is on concrete problems and examples.

We will showcase methods through detailed examples with associated code.
This presentation allows the academic level to be quite broad; understood by undergraduates all the way through advanced Ph.D. students.

X-13ARIMA-SEATS is one of, if not the most, widely used seasonal adjustment software within federal and statistical agencies.
Moreover, there is a movement in statistical agencies toward the use of R and open-source products.
This text is motivated to unify these two positions.
Additionally, it and also serve the following:

1.  Guide to professional seasonal adjustment with R To make the entry to the world of seasonal adjustment more accessible for those with an understanding of R.
    We leverage the users R knowledge to more easily understand the input/output of the X-13ARIMA-SEATS program.
    We also will give an overview of other possibilities of seasonal adjustment in R (e.g. stl, JDemetra).

2.  Focus on practitioner's problem To bridge an important gap in the training for many seasonal adjustment practitioners.
    The book addresses practical problems and shows how they can be addressed in X-13ARIMA-SEATS.
    The use of R allows them to have reproducible examples at hand.

## Overview of the book

## The [seasonalbook](https://github.com/christophsax/seasonalbook) package

An R package that supplements "Seasonal Adjustment in R", and contains all data and examples.

To install:

``` r
remotes::install_github("christophsax/seasonalbook")
```

Example series:

```{r}
library(seasonalbook)
plot(grocery)
```

Some Examples: Chinese New Year, structural breaks, direct or indirect seasonal adjustment, SEATS or X-11.
