## regARIMA Model

```{r}
#| results: "asis"
#| echo: false
source("_common.R")
status("polishing", "2022-12-21", section = "regARIMA Model")  # drafting, polishing, complete
```
<!-- - Idea of regARIMA -->

Depending on if you ultimately choose to perform a SEATS or X-11 seasonal adjustmnet, your usage of a regARIMA can differ. 
Both seasonal adjustmnet methods will use the regARIMA model to forecast the series in order to apply symmetric filters and extract components.
(but I thought SEATS was model-based? What do you mean SEATS uses filters? More on this in Chapter XXX).
SEATS will use the ARIMA model to derive components for the trend, seasonal and irregular terms.
The regARIMA model is also the place to include exogenous information or regressors into your seasonal adjusments.
Regressors such as holiday effects, additive outliers, levels shifts, or any user-defined effect can be included in your regARIMA model.
This can be thought of as solely a utility to forecast extend your series or a detection method to do inference.
In practice it is useful to use regARIMA modeling to answer questions such as, "Does my series have trading day effects?" or "Does this outlier effect my results?".
Another reason for regARIMA modeling is to include exogenous regression variables in our analysis.
These include, but are not limited to, moving holiday effects, trading day, outliers and level-shifts. 
The regARIMA modeling takes place prior to seasonal and can be iterative until a final model is reached. 
This section will focus on the left self-loop of our seasonal adjustment flowchat:

```{mermaid}
flowchart TD
  A[RegARIMA Modeling] --> B(Model Comparison / Diagnostics)
  B --> A
  A --> C[Seasonal Adjustment]
  C --> D[Seasonal Adjustment Diagnostics]
  D --> C
  D --> A
```

The regARIMA take the form
$$ f\left(\frac{Y_t}{D_t} \right) = \boldsymbol{\beta}^\prime {\mathbf X}_t + Z_t .$$
Here $Y_t$ is the observed time series. 
The function $f$ represents a transformation, most commonly used is the log transform ie $f(x) = \log(x)$.
$D_t$ is any intervention that has taken place prior to any transformation or modeling.
This intervention is usually subjective and customized for individual series on an as-needed basis.
For example, if a worker strike occured in a specific industry that effected certain economic series.
If no transformation or intervention is needed the model form is:
$$ Y_t = \underbrace{\boldsymbol{\beta}^\prime {\mathbf X}_t}_{\text{Regression}} + \underbrace{Z_t}_{\text{ARIMA}} .$$

The regression variables appear in the columns of the design matrix ${\mathbf X}_t$ and $Z_t$ is an ARIMA process.
This last assumption on $Z_t$ is what distinguished a regARIMA model from more classic linear models and multiple linear regression where error terms are assumed uncorrelated.

In order to achieve a suitable seasonal adjustment it is important to get the regARIMA model correct.
For most dataset the built in automatic modeling features of the X13 program will be suitable to detect a reasonable model. 
This can be used as a starting point for more rigorous regARIMA model development or used as the final regARIMA modeling choice for your seasonal adjustment needs.
We evoke automatic model identification through the XXX spec. 
The default behavior of the R seasonal package is XXX which includes automatic model identification. 


::: {.callout-tip}
## Automatic and manual model choice

As an aside, the general rule is to not use automatic modeling in production.
This mean, if you are going to include seasonal adjustment as part of a large scale data processing that occurs regularly (say monthly), then it is not advisable to have automatic model identification run every month.
Instead, an alternative process, is to run automodel once and then fix the model choice in the XXX spec file.
This does not need to be done manually since the `static()` function from the seasonal package can do this for you.
:::

| Outlier Type                | Automatic Detection Available? |
|-----------------------------|--------------------------------|
| Additive outliers (AO)      | Yes (default)                  |
| Level shifts (LS)           | Yes (default)                  |
| Temporary level shifts (TL) | Yes                            |
| Temporary changes (TC)      | No                             |
| Ramps (RP, QI, QD)          | No                             |
| Seasonal outliers (SO)      | No                             |

Consider the default seasonal adjustment:
```{r}
library(seasonal)
m <- seasonal::seas(AirPassengers, x11 = "")
print(m$spc$automdl)
print(m$spc$arima)
```
Notice the value `NULL` indicates no ARIMA model is specified and the returned arguments for the automdl spec indicate it is active during the X13 run. 

```{r}
seasonal::udg(m, "automdl")
```

Indicates that automatic modeling identified the (0 1 1)(0 1 1) model as the best choice. 
If we want to hardcode this model for subsequent runs, and turn off automatic model identification, this can be done via

```{r}
m_call <- seasonal::static(m)
m2 <- eval(m_call)
```
## Outlier 

There are many options you can modify when searching for outliers in your series.
Some of the most practical options to start your exploration are the *type*, *critical value* and *span* that you would like to search. 

Here is an example of using span to limit the outlier search to the last few years of a series:

```{r}
m_span <- seas(AirPassengers,
  outlier.types = c("ao", "ls", "tc"),
  outlier.critical = 4.0,
  outlier.span = "1958.jan, ")
summary(m_span)

m_nospan <- seas(AirPassengers,
  outlier.types = c("ao", "ls", "tc"),
  outlier.critical = 4.0)
summary(m_nospan)
```

The default critical value is XXX. 
Notice the MA-Nonseasonal-01 value when comparing `m_span` with `m_nospan`.
We see the choice of span, and ultimately the choise to include an outlier in your model can have a dramatic effect on the estimated regARIMA parameters.


## How to use the regression spec

## Table of all Regression effects available

 <!--
  (steal from CR slides)

-->


## Case study

Decide if you should include AO in May 2014.
Construct a simple user defined regressor to handle specific issue.

